### 2025-09-10 cosmology.tex
update particle cosmology

% Time-Stamp: <2022-05-16 15:23:42>

@@

\subsection{Particle cosmology}
The particle number density, pressure, and energy density are calculated from distribution functions:
\begin{alignat}{3}
 f\w{MB}(\vc k)&=\frac{g}{\ee^{(E-\mu)/T}},\qquad&
 f\w{BE}(\vc k)&=\frac{g}{\ee^{(E-\mu)/T}-1},\qquad&
 f\w{FD}(\vc k)&=\frac{g}{\ee^{(E-\mu)/T}+1};
\end{alignat}
\begin{alignat}{3}
 n   &=\int\ddP{k}f(\vc k),\qquad&
 \rho&=\int\ddP{k}Ef(\vc k),\qquad&
 p   &=\int\ddP{k}k_zv_z f(\vc k)
      =\int\ddP{k}\frac{k^2\cos^2\theta}{E} f(\vc k).
\end{alignat}
Note the pressure is (momentum)$\times$(flux per time) on a ``wall''; assuming MB, $p=\rho/3$ for $m\ll T$ and $p=T\rho/m$ for $m\gg T$.

A thermal average of a cross section $\sigma(s)$ is schematically given by
\begin{equation}
 \mean{\sigma v}_{AB\to12\cdots n}(T)
=\frac{1}{n_An_B}\int\ddP{k_A}\ddP{k_B}\left(
f_A f_B\right)\Bigl\{\phi_1 \cdots \phi_n\sigma(s)\Bigr\}
v_{\text{M\o l}};\qquad \phi_X = \ee^{(E-\mu)/T}f_X/g,
\end{equation}
Here, the final state statistical factor $\phi_1\cdots\phi_n$ are subject to the phase space integral of the calculation of $\sigma(s)$. They are specifically given by $\phi\w{MB}=1$, $\phi\w{BE}=1+f\w{BE}/g$, and $\phi\w{FD}=1-f\w{FD}/g$.
Similarly, a thermal averaged decay rate is given by
\begin{equation}
 \mean{\Gamma}_{A\to12\cdots n}
=\frac{1}{n_A}\int\ddP{k_A}
f_A \left\{\phi_1 \cdots \phi_n
\frac{m_A}{E_A}\Gamma\right\}.
\end{equation}

>>>>

% Time-Stamp: <2025-09-10 14:08:10>

@@

\subsection{Particle cosmology}
\subsubsection{Particle distribution and density}
In kinematic equilibrium, the particle number density, energy density, and pressure (i.e., (momentum)$\times$(flux per time) against a ``wall'') are calculated from distributions
\begin{alignat}{3}
 f\w{MB}(\vc k)&=\frac{g}{\ee^{(E-\mu)/T}},\qquad&
 f\w{BE}(\vc k)&=\frac{g}{\ee^{(E-\mu)/T}-1},\qquad&
 f\w{FD}(\vc k)&=\frac{g}{\ee^{(E-\mu)/T}+1},
\end{alignat}
\begin{alignat}{3}
 n   &=\int\ddP{k}f(\vc k),\qquad&
 \rho&=\int\ddP{k}Ef(\vc k),\qquad&
 p   &=\int\ddP{k}k_zv_z f(\vc k)
      =\int\ddP{k}\frac{k^2\cos^2\theta}{E} f(\vc k),
\end{alignat}
where we, in principle, take $\mu<m$ for Bose--Einstein statistics.
It is also useful to define the difference of particle and antiparticle number densities, assuming their chemical potential is opposite in sign:
\begin{equation}
  n_\Delta \deq n_+ - n_- = \int\ddP{k}\Bigl[f(\vc k,\mu) - f(\vc k,-\mu)\Bigr].
\end{equation}

For Maxwell--Boltzmann (MB) distribution, they are given explicitly by, with $\nu\deq \mu/T$ and $x\deq m/T$,
\begin{equation}
  n\w{MB} = \frac{gT^3\ee^{\nu}}{2\pi^2}x^2K_2(x),\quad
  %n\w{\Delta;MB} = g\sinh{\nu}\cdot\frac{T^3}{\pi^2}x^2K_2(x),\quad
  \rho\w{MB} = \left(3+\frac{xK_1(x)}{K_2(x)}\right)T n\w{MB},\quad
  p\w{MB} = T n\w{MB},
\end{equation}
which reduce in special cases to
\begin{align}
  &\text{[massless: $m\ll T$]}
  &&n\w{MB} = \frac{gT^3\ee^{\nu}}{\pi^2}\left(1-\frac14 x^2+\Order(x^4)\right),\quad
  \rho\w{MB} = 3Tn\w{MB}\left(1+\frac16x^2+\Order(x^4)\right),
\\&\text{[low-temperature: $m\gg T$]}
  &&n\w{MB} \approx \ee^{(\mu-m)/T}\left({\frac{m T}{2\pi}}\right)^{3/2},\quad
  \rho\w{MB} \approx \left(m+\frac32T\right)n\w{MB}.
\end{align}

\pagebreak

For Fermi--Dirac or Bose--Einstein distributions, the results are not expressed analytically, even with $\mu=0$. Namely,
\begin{equation}
  n=\int_m^\infty\frac{\dd E}{2\pi^2}\frac{gE\sqrt{E^2-m^2}}{\ee^{(E-\mu)/T}\pm1},
  \quad
  \rho=\int_m^\infty\frac{\dd E}{2\pi^2}\frac{gE^2\sqrt{E^2-m^2}}{\ee^{(E-\mu)/T}\pm1},
  \quad
  p=\frac13\int_m^\infty\frac{\dd E}{2\pi^2}\frac{g(E^2-m^2)^{3/2}}{\ee^{(E-\mu)/T}\pm1}.
\end{equation}
Analytical expressions are available for the massless limit as
\begin{align}
    &\text{[massless: $m\ll T$]}
  &&
  n\w{FD}    = \frac{g L_3^-}{\pi^2}+\Order(x),\quad
  \rho\w{FD} = \frac{3g L_4^-}{\pi^2}+\Order(x^2),\quad
  p\w{FD}    = \frac{g L_4^-}{\pi^2}+\Order(x^2),\\
   &\text{[massless: $m\ll T$; $\mu<0$]}
  &&
  n\w{BE}    = \frac{g L_3^+}{\pi^2}+\Order(x),\quad
  \rho\w{BE} = \frac{3g L_4^+}{\pi^2}+\Order(x^2),\quad
  p\w{BE}    = \frac{g L_4^+}{\pi^2}+\Order(x^2),
\end{align}
where $L^\pm_n=\pm\mathop{\text{\texttt{PolyLog}}}(n,\pm\ee^{\mu/T})$.
Note we may need thermal masses for proper discussion, especially for the Bose--Einstein case, which is well-defined only for $\mu<m$.

The difference $n_\Delta$ can be expressed under the assumption $\mu/T\ll1$:
\begin{align}
 &\text{[near equilibrium: $\mu\ll T$]}
  &&
 (n_\Delta{})^{\mathrm{BE}}_{\mathrm{FD}}
 =\int_{m}^\infty
 \frac{\dd E}{\pi^2} \frac{gE\sqrt{E^2-m^2}\ee^{E/T}}{(\ee^{E/T}\mp1)^2}\frac{\mu}{T}
 +\Order\left(\frac{\mu^3}{T^3}\right)
 =g\mu T^2\xi_\mp(x)
 +\Order\left(\frac{\mu^3}{T^3}\right),
\end{align}
where $\displaystyle
 \xi_\mp(x)\deq \int_{x}^\infty\frac{k\ee^{k}\sqrt{k^2-x^2}\,\dd k}{(\ee^{k}\mp1)^2\pi^2}
$ can be calculated numerically and remain finite.
For fermions,
\begin{align}
 \text{[$\mu\ll T$, $m\ll T$]}\qquad
 (n_\Delta{})_{\mathrm{FD}} = g\mu\left[\frac{T^2}{6} - \frac{m^2}{4\pi^2}
 +\Order\left(\frac{m^4}{T^4}\right)\right]
 +\Order\left(\frac{\mu^2}{T^2}\right).
\end{align}
For Bose--Einstein case, it results in $\xi_-(0)=1/3$ and hence $n_{\Delta;\mathrm{BE}}\questeq g\mu T^2/3$ with no divergence, but we may discuss it is invalid because of the implicit assumption $\mu<m$.

\subsubsection{Thermal average of cross section and decay rate}
A thermal average of a cross section $\sigma(s)$ is schematically given by
\begin{equation}
 \mean{\sigma v}_{AB\to12\cdots n}(T)
=\frac{1}{n_An_B}\int\ddP{k_A}\ddP{k_B}\left(
f_A f_B\right)\Bigl\{\phi_1 \cdots \phi_n\sigma(s)\Bigr\}
v_{\text{M\o l}};\qquad \phi_X = \ee^{(E-\mu)/T}f_X/g,
\end{equation}
Here, the final state statistical factor $\phi_1\cdots\phi_n$ are subject to the phase space integral of the calculation of $\sigma(s)$. They are specifically given by $\phi\w{MB}=1$, $\phi\w{BE}=1+f\w{BE}/g$, and $\phi\w{FD}=1-f\w{FD}/g$.
Similarly, a thermal averaged decay rate is given by
\begin{equation}
 \mean{\Gamma}_{A\to12\cdots n}
=\frac{1}{n_A}\int\ddP{k_A}
f_A \left\{\phi_1 \cdots \phi_n
\frac{m_A}{E_A}\Gamma\right\}.
\end{equation}
